Deep mutational scanning of hemagglutinin helps distinguish the evolutionary fate of human H3N2 influenza virus lineages

### 2018-01-16 paper/manuscript.tex
Add new WSN logoplot to supplementary figures
Full draft of Discussion

\begin{abstract}
% no more than 250 words. (199 now)
Human influenza virus rapidly accumulates mutations in its major surface protein hemagglutinin (HA).
Forecasting the evolutionary success of influenza virus strains requires understanding how these mutations affect the protein's functionality and antigenicity.
Here we experimentally measure the effects on viral growth in cell culture of all single amino-acid mutations to the HA from a recent human H3N2 influenza virus strain. 
We show that these measurements are informative for understanding viral evolution in nature, since mutations in evolutionarily successful H3N2 viral lineages are measured to generally be more favorable than mutations in lineages that die out.
We also compare our measurements to prior data for an H1 HA, and find that the amino-acid preferences of many sites have substantially shifted between H1 and H3 HAs.
For instance, there is much less of a disparity in mutational tolerance between the head and stalk domains for the H3 HA.
Overall, our work suggests that experimental measurements of mutational effects might be leveraged to help forecast the evolutionary fates of viral lineages in nature.
\end{abstract}

@@

\section*{Discussion}
\label{sec:discussion}
We have measured the effect of all possible single amino-acid mutations to Perth/2009 H3 on viral growth in cell culture, and shown that these measurements are informative about the evolutionary fate of human H3N2 viral strains in nature.
Specifically, strains the trunk of the phylogenetic tree tend to have mutations that that our experiments score as more favorable than mutations on the side branches.
Distinguishing between the trunk and side branches is the equivalent of identifying which strains succeed in nature, and so the ability of our experiments to do this show that they are useful.

It may seem surprising that simple measurements of viral growth in cell culture are informative about fitness in nature.
There are many examples of mutations that have different effects in the lab and in nature, and our experiments definitely don't measure all the relevant factors.
In addition, we only look at mutations to HA.
Nonetheless, prior to our work there was no comprehensive estimates the functional effects of mutations on viral fitness outside of antigenicity.
Therefore, all prior quantitative modeling either neglected non-antigenic effects or simply assumed that all mutations outside epitope sites had an equivalent deleterious effect on viral fitness.
While our measurements are certainly imperfect, they are better than this.
... also suggests that measurements in more realistic experimental models should be even more useful.

Another caveat is that we are making measurements of the effects of mutations in a single HA variant, Perth/2009.
We are then extrapolating those measurements across the full spectrum of seasonal human H3N2.
These extrapolations will only be valid to the extent that the effects of mutations remain conserved during HA's evolution.
It is known from extensive work on protein evolution that mutational effects can shift due to epistasis, and the extent to which this occurs is an important question in its own right.
The success of our measurements across all H3N2 show that these shifts aren't too huge for relatively similar strains.
But we also compared our measurements for H3 to prior measurements on H1 and found that there were large shifts --- much greater than those observed in prior protein-wide comparisons of more closely related homologs.
The effective use of experiments for studying evolution in nature will require further analysis of how mutational effects shift in order to figure out how widely the results of any given experiment can be applied.

Other things are important too.
We did not look at antigenicity (whereas prior studies have only looked at that).
In the future, our experimental approach could be used to look at antigenicity too.

@@

\begin{suppfigure}
\centerline{\includegraphics[width=\textwidth]{figs/S03_WSNprefs_logoplot/WSN-rescaled_prefs.pdf}}
\caption{\label{suppfig:WSNprefs_logoplot}
{\bf The site-specific amino-acid preferences of the WSN/1933 H1 HA.} 
The amino-acid preferences from~\cite{doud2016accurate} after taking the average of the experimental replicates and re-scaling~\cite{hilton2017phydms} by the stringency parameter \comment{somehow indicate where this stringency parameter}.
The sites are in H3 numbering.
\comment{I would suggest just getting rid of RSA. You might be able to replace the next two sentences by just saying: the overlays show the same information as in Figure X}.
The bottom overlay bar indicates the HA domain (sig. pep. = signal peptide, HA1 ecto. = HA1 ectodomain, HA2 ecto. = HA2 ectodomain, TM = transmembrane domain, cyto. tail. = cytoplasmic tail).
The letters directly above each logo stack indicate the wildtype amino acid at that site.
}
\end{suppfigure}

>>>>

\begin{abstract}
% no more than 250 words. (170 now)
Human influenza virus rapidly accumulates mutations in its major surface protein hemagglutinin (HA).
Forecasting the evolutionary success of influenza virus strains requires understanding how these mutations affect the protein's functionality and antigenicity.
Here we experimentally measure the effects on viral growth in cell culture of all single amino-acid mutations to the HA from a recent human H3N2 influenza virus strain. 
We show that these measurements are informative for understanding viral evolution in nature, since mutations in evolutionarily successful H3N2 viral lineages are measured to generally be more favorable than mutations in lineages that die out.
We also compare our measurements to prior data for an H1 HA, and find that the amino-acid preferences of many sites have substantially shifted between H1 and H3 HAs.
For instance, there is much less of a disparity in mutational tolerance between the head and stalk domains for the H3 HA.
Overall, our work suggests that experimental measurements of mutational effects might be leveraged to help forecast the evolutionary fates of viral lineages in nature.
\end{abstract}

@@

\section*{Discussion}
\label{sec:discussion}
We have measured the effect of all possible single amino-acid mutations to the Perth/2009 H3 on viral growth in cell culture, and demonstrated that these measurements can be informative for distinguishing the evolutionary fate of human H3N2 viral strains in nature.
Specifically, strains on the trunk of the phylogenetic tree tend to have mutations that are beneficial for viral growth in our experiments when compared to side-branch mutations.
Differentiating between trunk and side branch lineages can enable forecasting of influenza virus evolution by identifying which strains are likely to succeed in nature.
The fact that our experiments are able to reveal differences between trunk and side branch lineages highlight the potential for incorporating experimental measurements of the non-antigenic effects of mutations for evolutionary forecasting.

Surprisingly, our simple experimental measurements of viral growth in cell culture helped to elucidate differences in fitness of viral strains in nature.
Mutations may have effects in the laboratory that are distinct from their effects in nature~\cite{daniels1985fusion,sun2010modifications,lee2013comparison,wu2017structural}, and our experiments certainly don't capture all of the relevant selective pressures on HA that exist in nature.
Mutations in viral genes other than HA are also important in determining strain success~\cite{memoli2009recent,raghwani2017selection}.
Yet, prior to our work presented here, there were no comprehensive studies of the functional effects of mutations to HA on viral fitness outside of antigenicity.
Previous quantitative modeling work either omitted the non-antigenic effects of mutations~\cite{sun2013using,harvey2016identification,neher2016prediction} or made the simplifying assumption that all mutations outside epitope sites had equivalent deleterious effects on viral fitness~\cite{luksza2014predictive}.
While our measurements may be imperfect, we have demonstrated that these measurements are still of utility in distinguishing the selective advantage of successful influenza viral lineages over $\sim$50 years of evolution.
This also suggests that measurements in experimental models that more faithfully recapitulate the selective forces in nature should be of even greater utility for differentiating evolutionarily successful viral lineages.

In addition, we are measuring the effects of mutations in a single HA variant from the Perth/2009 strain and extrapolating these measurements to other seasonal human H3N2 strains across a 50-year timespan.
However, these extrapolations will only be valid to the extent that the effects of mutations remain conserved during HA's evolution.
Extensive work on protein evolution has shown that epistasis can shift the effects of mutations~\cite{gong2013stability,natarajan2013epistasis,harms2014historical,starr2016epistasis,starr2017alternative,haddox2017mapping}.
% and the extent to which this occurs is an important question in its own right.
The observation that our measurements score trunk mutations more favorably than side branch mutations across most of H3N2 evolution suggests that for relatively closely related strains, the mutational effects are not shifted to a large extent.
On the other hand, when we compared our measurements for H3 to prior measurements on H1, we found substantial shifts at many sites --- much greater than those observed in prior protein-wide comparisons of more closely related homologs.
Further analysis of how mutational effects shift as protein variants diverge can provide insight into how broadly the results of a given experiment can be generalized across variants.

Importantly, the antigenic properties of HA are integral in determining strain success, but we did not study the effect of mutations on antigenicity.
However, our high-throughput experimental approach can be harnessed to profile mutations that confer escape from antibodies, as has been performed for the WSN/1933 H1 HA~\cite{doud2017complete,doud2017quantifying} and HIV Env~\cite{dingens2017comprehensive}.
In future, these antigenic studies can be combined with functional measurements to inform forecasting of the evolution of seasonal influenza virus.

@@

\begin{suppfigure}
\centerline{\includegraphics[width=\textwidth]{figs/S03_WSNprefs_logoplot/WSN-rescaled_prefs.pdf}}
\caption{\label{suppfig:WSNprefs_logoplot}
{\bf The site-specific amino-acid preferences of the WSN/1933 H1 HA.} 
The amino-acid preferences from~\cite{doud2016accurate} after taking the average of the experimental replicates and re-scaling~\cite{hilton2017phydms} by the stringency parameter \comment{somehow indicate where this stringency parameter}.
The sites are in H3 numbering.
The overlays show the same information as in Figure~\ref{fig:logoplot} (domain and wildtype amino acid).
}
\end{suppfigure}
